### 2014-10-13 doc/thesis/system.tex
Adding integer to formal type system
Fixing the type checking of multiple assignments

\begin{figure}[!ht]
\textbf{Subtyping}\\
\dstart
\begin{footnotesize}
$$
\begin{array}{c}
\begin{array}{c}
\mylabel{S-LITERAL}\\
\senv \vdash L \subtype L
\end{array}
\;
\begin{array}{c}
\mylabel{S-FALSE}\\
\senv \vdash \False \subtype \Boolean
\end{array}
\;
\begin{array}{c}
\mylabel{S-TRUE}\\
\senv \vdash \True \subtype \Boolean
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-NUMBER}\\
\senv \vdash {\it number} \subtype \Number
\end{array}
\;
\begin{array}{c}
\mylabel{S-STRING}\\
\senv \vdash {\it string} \subtype \String
\end{array}
\;
\begin{array}{c}
\mylabel{S-BASE}\\
\senv \vdash B \subtype B
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-NIL}\\
\senv \vdash \Nil \subtype \Nil
\end{array}
\;
\begin{array}{c}
\mylabel{S-VALUE}\\
\senv \vdash T \subtype \Value
\end{array}
\;
\begin{array}{c}
\mylabel{S-ANY}\\
\senv \vdash \Any \subtype \Any
\end{array}
\;
\begin{array}{c}
\mylabel{S-SELF}\\
\senv \vdash \Self \subtype \Self
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-UNION1}\\
\dfrac{\senv \vdash T_{1} \subtype T \;\;\;
       \senv \vdash T_{2} \subtype T}
      {\senv \vdash T_{1} \cup T_{2} \subtype T}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UNION2}\\
\dfrac{\senv \vdash T \subtype T_{1}}
      {\senv \vdash T \subtype T_{1} \cup T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UNION3}\\
\dfrac{\senv \vdash T \subtype T_{2}}
      {\senv \vdash T \subtype T_{1} \cup T_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FUNCTION}\\
\dfrac{\senv \vdash P_{2} \subtype P_{1} \;\;\;
       \senv \vdash R_{1} \subtype R_{2}}
      {\senv \vdash P_{1} \rightarrow R_{1} \subtype P_{2} \rightarrow R_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-TABLE1}\\
\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
       \senv \vdash F_{j} \subtype F_{i}}
      {\senv \vdash \{F_{1}, ..., F_{m}\}_{c} \subtype \{F_{1}, ..., F_{n}\}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-TABLE2}\\
\dfrac{\forall i \in 1..m \; \senv \vdash F_{i}' \subtype F_{i}}
      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \subtype \{F_{1}', ..., F_{m}', ..., F_{n}'\}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-TABLE3}\\
\dfrac{\forall i \in 1..m \; \senv \vdash F \subtype F_{i}}
      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \subtype \{F\}} 
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD1}\\
\dfrac{\senv \vdash K_{2} \subtype K_{1} \;\;\;
       \senv \vdash T_{1} \subtype T_{2} \;\;\;
       \senv \vdash T_{2} \subtype T_{1}}
      {\senv \vdash K_{1}:T_{1} \subtype K_{2}:T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD2}\\
\dfrac{\senv \vdash K_{2} \subtype K_{1} \;\;\;
       \senv \vdash T_{1} \subtype T_{2}}
      {\senv \vdash \Const \; K_{1}:T_{1} \subtype \Const \; K_{2}:T_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD3}\\
\dfrac{\senv \vdash K_{2} \subtype K_{1} \;\;\;
       \senv \vdash T_{1} \subtype T_{2}}
      {\senv \vdash K_{1}:T_{1} \subtype \Const \; K_{2}:T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-VARIABLE}\\
\dfrac{X_{1} \subtype X_{2} \in \senv}
      {\senv \vdash X_{1} \subtype X_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-RECURSIVE1}\\
\dfrac{\senv, X_{1} \subtype X_{2} \vdash T_{1} \subtype T_{2}}
      {\senv \vdash \mu X_{1}.T_{1} \subtype \mu X_{2}.T_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-RECURSIVE2}\\
\dfrac{X \subtype X \not\in \senv \;\;\;
      \senv, X \subtype X \vdash T_{1} \subtype T_{2}}
      {\senv \vdash \mu X.T_{1} \subtype T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-RECURSIVE3}\\
\dfrac{X \subtype X \in \senv \;\;\;
       \senv \vdash X \subtype T_{2}}
      {\senv \vdash \mu X.T_{1} \subtype T_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-RECURSIVE4}\\
\dfrac{X \subtype X \not\in \senv \;\;\;
       \senv, X \subtype X \vdash T_{1} \subtype T_{2}}
      {\senv \vdash T_{1} \subtype \mu X.T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-RECURSIVE5}\\
\dfrac{X \subtype X \in \senv \;\;\;
       \senv \vdash T_{1} \subtype X}
      {\senv \vdash T_{1} \subtype \mu X.T_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-VOID}\\
\senv \vdash \Void \subtype \Void
\end{array}
\;
\begin{array}{c}
\mylabel{S-VARARG}\\
\dfrac{\senv \vdash T_{1} \cup \Nil \subtype T_{2} \cup \Nil}
      {\senv \vdash T_{1}* \subtype T_{2}*}
\end{array}
\;
\begin{array}{c}
\mylabel{S-TUPLE1}\\
\dfrac{\senv \vdash T_{1} \subtype T_{2} \;\;\;
       \senv \vdash S_{1} \subtype S_{2}}
      {\senv \vdash T_{1} \times S_{1} \subtype T_{2} \times S_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-TUPLE2}\\
\dfrac{\senv \vdash T_{1} \cup \Nil \subtype T_{2} \;\;\;
       \senv \vdash T_{1}* \subtype S_{2}}
      {\senv \vdash T_{1}* \subtype T_{2} \times S_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-TUPLE3}\\
\dfrac{\senv \vdash T_{1} \subtype T_{2} \cup \Nil \;\;\;
       \senv \vdash S_{1} \subtype T_{2}*}
      {\senv \vdash T_{1} \times S_{1} \subtype T_{2}*}
\end{array}
\end{array}
$$
\end{footnotesize}
\dend
\caption{The subtyping relation}
\label{fig:subtyping}
\end{figure}

\begin{figure}[!ht]
\textbf{Consistent-Subtyping}\\
\dstart
\begin{footnotesize}
$$
\begin{array}{c}
\begin{array}{c}
\mylabel{C-LITERAL}\\
\senv \vdash L \lesssim L
\end{array}
\;
\begin{array}{c}
\mylabel{C-FALSE}\\
\senv \vdash \False \lesssim \Boolean
\end{array}
\;
\begin{array}{c}
\mylabel{C-TRUE}\\
\senv \vdash \True \lesssim \Boolean
\end{array}
\\ \\
\begin{array}{c}
\mylabel{C-NUMBER}\\
\senv \vdash {\it number} \lesssim \Number
\end{array}
\;
\begin{array}{c}
\mylabel{C-STRING}\\
\senv \vdash {\it string} \lesssim \String
\end{array}
\;
\begin{array}{c}
\mylabel{C-BASE}\\
\senv \vdash B \lesssim B
\end{array}
\\ \\
\begin{array}{c}
\mylabel{C-NIL}\\
\senv \vdash \Nil \lesssim \Nil
\end{array}
\;
\begin{array}{c}
\mylabel{C-VALUE}\\
\senv \vdash T \lesssim \Value
\end{array}
\;
\begin{array}{c}
\mylabel{C-ANY1}\\
\senv \vdash T \lesssim \Any
\end{array}
\;
\begin{array}{c}
\mylabel{C-ANY2}\\
\senv \vdash \Any \lesssim T
\end{array}
\;
\begin{array}{c}
\mylabel{C-SELF}\\
\senv \vdash \Self \lesssim \Self
\end{array}
\\ \\
\begin{array}{c}
\mylabel{C-UNION1}\\
\dfrac{\senv \vdash T_{1} \lesssim T \;\;\;
       \senv \vdash T_{2} \lesssim T}
      {\senv \vdash T_{1} \cup T_{2} \lesssim T}
\end{array}
\;
\begin{array}{c}
\mylabel{C-UNION2}\\
\dfrac{\senv \vdash T \lesssim T_{1}}
      {\senv \vdash T \lesssim T_{1} \cup T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{C-UNION3}\\
\dfrac{\senv \vdash T \lesssim T_{2}}
      {\senv \vdash T \lesssim T_{1} \cup T_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{C-FUNCTION}\\
\dfrac{\senv \vdash P_{2} \lesssim P_{1} \;\;\;
       \senv \vdash R_{1} \lesssim R_{2}}
      {\senv \vdash P_{1} \rightarrow R_{1} \lesssim P_{2} \rightarrow R_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{C-TABLE1}\\
\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
       \senv \vdash F_{j} \lesssim F_{i}}
      {\senv \vdash \{F_{1}, ..., F_{m}\}_{c} \lesssim \{F_{1}, ..., F_{n}\}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{C-TABLE2}\\
\dfrac{\forall i \in 1..m \; \senv \vdash F_{i}' \lesssim F_{i}}
      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \lesssim \{F_{1}', ..., F_{m}', ..., F_{n}'\}}
\end{array}
\;
\begin{array}{c}
\mylabel{C-TABLE3}\\
\dfrac{\forall i \in 1..m \; \senv \vdash F \lesssim F_{i}}
      {\senv \vdash \{F_{1}, ..., F_{m}\}_{o} \lesssim \{F\}} 
\end{array}
\\ \\
\begin{array}{c}
\mylabel{C-FIELD1}\\
\dfrac{\senv \vdash K_{2} \lesssim K_{1} \;\;\;
       \senv \vdash T_{1} \lesssim T_{2} \;\;\;
       \senv \vdash T_{2} \lesssim T_{1}}
      {\senv \vdash K_{1}:T_{1} \lesssim K_{2}:T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{C-FIELD2}\\
\dfrac{\senv \vdash K_{2} \lesssim K_{1} \;\;\;
       \senv \vdash T_{1} \lesssim T_{2}}
      {\senv \vdash \Const \; K_{1}:T_{1} \lesssim \Const \; K_{2}:T_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{C-FIELD3}\\
\dfrac{\senv \vdash K_{2} \lesssim K_{1} \;\;\;
       \senv \vdash T_{1} \lesssim T_{2}}
      {\senv \vdash K_{1}:T_{1} \lesssim \Const \; K_{2}:T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{C-VARIABLE}\\
\dfrac{X_{1} \lesssim X_{2} \in \senv}
      {\senv \vdash X_{1} \lesssim X_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{C-RECURSIVE1}\\
\dfrac{\senv, X_{1} \lesssim X_{2} \vdash T_{1} \lesssim T_{2}}
      {\senv \vdash \mu X_{1}.T_{1} \lesssim \mu X_{2}.T_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{C-RECURSIVE2}\\
\dfrac{X \lesssim X \not\in \senv \;\;\;
      \senv, X \lesssim X \vdash T_{1} \lesssim T_{2}}
      {\senv \vdash \mu X.T_{1} \lesssim T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{C-RECURSIVE3}\\
\dfrac{X \lesssim X \in \senv \;\;\;
       \senv \vdash X \lesssim T_{2}}
      {\senv \vdash \mu X.T_{1} \lesssim T_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{C-RECURSIVE4}\\
\dfrac{X \lesssim X \not\in \senv \;\;\;
       \senv, X \lesssim X \vdash T_{1} \lesssim T_{2}}
      {\senv \vdash T_{1} \lesssim \mu X.T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{C-RECURSIVE5}\\
\dfrac{X \lesssim X \in \senv \;\;\;
       \senv \vdash T_{1} \lesssim X}
      {\senv \vdash T_{1} \lesssim \mu X.T_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{C-VOID}\\
\senv \vdash \Void \lesssim \Void
\end{array}
\;
\begin{array}{c}
\mylabel{C-VARARG}\\
\dfrac{\senv \vdash T_{1} \cup \Nil \lesssim T_{2} \cup \Nil}
      {\senv \vdash T_{1}* \lesssim T_{2}*}
\end{array}
\;
\begin{array}{c}
\mylabel{C-TUPLE1}\\
\dfrac{\senv \vdash T_{1} \lesssim T_{2} \;\;\;
       \senv \vdash S_{1} \lesssim S_{2}}
      {\senv \vdash T_{1} \times S_{1} \lesssim T_{2} \times S_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{C-TUPLE2}\\
\dfrac{\senv \vdash T_{1} \cup \Nil \lesssim T_{2} \;\;\;
       \senv \vdash T_{1}* \lesssim S_{2}}
      {\senv \vdash T_{1}* \lesssim T_{2} \times S_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{C-TUPLE3}\\
\dfrac{\senv \vdash T_{1} \lesssim T_{2} \cup \Nil \;\;\;
       \senv \vdash S_{1} \lesssim T_{2}*}
      {\senv \vdash T_{1} \times S_{1} \lesssim T_{2}*}
\end{array}
\end{array}
$$
\end{footnotesize}
\dend
\caption{The consistent-subtyping relation}
\label{fig:consistent_subtyping}
\end{figure}

>>>>

In our type system function types have the form $p \rightarrow r$,
where $p$ is the type of the parameters list and $r$ is the
return type.
The type of the parameters list can be $\Void$ or a tuple type,
while the return type can be $\Void$, a tuple type, or a union of tuples.
The type $\Void$ means an empty tuple.
Our type system includes union of tuples on the return type because
in Lua it is common to overload the return type to denote error,
as we mentioned on Section \ref{sec:statistics}.

The type $\Void$ and union of tuple types are also second-level types.
The type $\Void$ is subtype only of itself.
The subtyping rules for union of tuple types are similar to the
subtyping rules for union of first-level types.
Next we show the subtyping rules for these second-level types:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-VOID}\\
\senv \vdash \Void \subtype \Void
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-UNION4}\\
\dfrac{\senv \vdash s_{1} \subtype s \;\;\;
       \senv \vdash s_{2} \subtype s}
      {\senv \vdash s_{1} \sqcup s_{2} \subtype s}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UNION5}\\
\dfrac{\senv \vdash s \subtype s_{1}}
      {\senv \vdash s \subtype s_{1} \sqcup s_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UNION6}\\
\dfrac{\senv \vdash s \subtype s_{2}}
      {\senv \vdash s \subtype s_{1} \sqcup s_{2}}
\end{array}
\end{array}
\]

The subtyping rule for function types is standard, that is, it is
contravariant on the type of the parameters list and
covariant on the return type:
\[
\begin{array}{c}
\mylabel{S-FUNCTION}\\
\dfrac{\senv \vdash p_{2} \subtype p_{1} \;\;\;
       \senv \vdash r_{1} \subtype r_{2}}
      {\senv \vdash p_{1} \rightarrow r_{1} \subtype p_{2} \rightarrow r_{2}}
\end{array}
\]

\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-TABLE1}\\
\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
       \senv \vdash f_{j} \subtype f_{i}}
      {\senv \vdash \{f_{1}, ..., f_{m}\}_{c} \subtype \{f_{1}, ..., f_{n}\}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-TABLE2}\\
\dfrac{\forall i \in 1..m \; \senv \vdash f_{i}' \subtype f_{i}}
      {\senv \vdash \{f_{1}, ..., f_{m}\}_{o} \subtype \{f_{1}', ..., f_{m}', ..., f_{n}'\}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-TABLE3}\\
\dfrac{\forall i \in 1..m \; \senv \vdash f \subtype f_{i}}
      {\senv \vdash \{f_{1}, ..., f_{m}\}_{o} \subtype \{f\}} 
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD1}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2} \;\;\;
       \senv \vdash t_{2} \subtype t_{1}}
      {\senv \vdash k_{1}:t_{1} \subtype k_{2}:t_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD2}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2}}
      {\senv \vdash \Const \; k_{1}:t_{1} \subtype \Const \; k_{2}:t_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD3}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2}}
      {\senv \vdash k_{1}:t_{1} \subtype \Const \; k_{2}:t_{2}}
\end{array}
\end{array}
\]

Our type system also includes recursive types: $\mu x.t$,
where $t$ is a function of table type with $x$ appearing anywhere
a function of table type could appear.
For instance, $\mu x.\{``info":\Number, ``next":x \cup \Nil\}$
represents the type of single linked lists of numbers.
We introduce the following subtyping rules for recursive types:

\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-VARIABLE}\\
\dfrac{x_{1} \subtype x_{2} \in \senv}
      {\senv \vdash x_{1} \subtype x_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-AMBER}\\
\dfrac{\senv, x_{1} \subtype x_{2} \vdash t_{1} \subtype t_{2}}
      {\senv \vdash \mu x_{1}.t_{1} \subtype \mu x_{2}.t_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-AMBERL}\\
\dfrac{\senv \vdash t_{1}[x \mapsto \mu x.t_{1}] \subtype t_{2}}
      {\senv \vdash \mu x.t_{1} \subtype t_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-AMBERR}\\
\dfrac{\senv \vdash t_{1} \subtype t_{2}[x \mapsto \mu x.t_{2}]}
      {\senv \vdash t_{1} \subtype \mu x.t_{2}}
\end{array}
\end{array}
\]

Our type system also introduces projection types.
We shall see in the next section that our type system uses
projection types to handle union of typle types when they
appear on the right-hand side of the declaration of local variables.
For this reason, they do not have subtyping rules.

\[
\begin{array}{c}
\begin{array}{c}
\mylabel{C-ANY1}\\
\senv \vdash t \lesssim \Any
\end{array}
\;
\begin{array}{c}
\mylabel{C-ANY2}\\
\senv \vdash \Any \lesssim t
\end{array}
\end{array}
\]
